\textbf{ Analytic Light Curves in Reflected Light: \\% Phase Curves, Occultations, and Non-Lambertian Scattering for Spherical Planets and Moons }

### 2021-02-25 tex/ms.tex
acknowledgm

The present paper is an extension to the \starry algorithm,
adapting it to model phase curves and occultation light curves
in reflected light. The expressions
we derive are analytic: they may all be expressed in closed form
in terms of algebraic operations involving trigonometric functions
and (at times) elliptic integrals. We derive numerically
stable recursion relations for the efficient evaluation of all
expressions and code them within an autodifferentiation framework
to enable the computation of accurate derivatives for use in
gradient-based inference and optimization schemes.
Our code is fully
\href{https://github.com/rodluger/starry}{open-source},
comprehensively
\href{https://dev.azure.com/rodluger/starry/_test/analytics?definitionId=4}{unit-tested},
and supplemented with an extensive
\href{https://rodluger.github.io/starry}{API documentation}
and suite of
\href{https://rodluger.github.io/starry/latest/tutorials.html}{tutorials}.
%
As in \citet{Luger2019}, in the caption of each of the figures
we provide links \codeicon\ to the exact \Python scripts that
generated them. Next to many of the equations we also provide
links \prooficon\ to \textsf{Jupyter} notebooks containing
detailed derivations and/or validations.

@@

The algorithm presented in this paper has been implemented in the \Python
package \starry, which can be installed from
\href{https://github.com/rodluger/starry}{GitHub} or via the Python package
manager \textsf{pip}. The algorithm is coded in a mixture of \textsf{C++}
with forward automatic differentiation using the \textsf{Eigen} library
\citep{eigen} and \Python with backward differentiation using
just-in-time compiled \textsf{theano}
operations \citep{theano}. The user interface, however, is purely in
\Python. The \textsf{theano} backend facilitates integration with the
\textsf{exoplanet} modeling package \citep{exoplanet} and in particular
with \textsf{pymc3} \citep{pymc3} for inference with gradient-based
Markov Chain Monte Carlo (MCMC) schemes
such as Hamiltonian Monte Carlo \citep[HMC;][]{Duane1987}
and No-U-Turn Sampling \citep[NUTS;][]{Hoffman2011}.
%
Complete \href{https://rodluger.github.io/starry}{documentation} and an
extensive library of tutorials is available online.
The links next to each
of the figures \codeicon\, point to the \Python scripts used to generate
them and may also help in learning how to use \starry.

@@

It is important to note, however, that the \starry model is linear
only in the map coefficients $\mathbf{y}$. In any real application,
there will be uncertainty in the inputs of $\mathbf{X}$, such
as the orbital parameters, the occultor parameters, the scattering
law, etc. These parameters must typically be sampled over, since the model
is a nonlinear function of them. However, the analyticity---and in particular,
the differentiability---of the \starry model makes sampling via
gradient-based MCMC easy. Moreover, the linearity of the model
with respect to $\mathbf{y}$ allows one to efficiently marginalize over
those parameters when sampling over the nonlinear parameters.
Tutorials on how to do this can be found in the
\starry \href{https://rodluger.github.io/starry}{documentation}.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_i \neq \vartheta_r$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.  Lastly, the presence
of time-variable features, such as clouds, dust storms, or seasonal
variations of vegetation are not accounted for in the model, although
the \href{https://rodluger.github.io/starry}{documentation} discusses
how one may approach the modeling of temporal features.

Finally, we would like to emphasize that while spherical harmonics are
a convenient basis for the purpose of computing light curves, they
have a significant drawback when it comes to modeling real planetary
surfaces: it can be difficult to strictly enforce physical
values of the albedo everywhere on the surface map when doing inference.
That is because there is no analytic way to determine whether a
spherical harmonic representation is positive-valued
(or restricted to a given range) everywhere on the sphere.
Instead, this must be checked numerically, by evaluating the function on a
discrete grid. This makes it somewhat cumbersome to implement positivity as a prior
when doing inference; in particular, this prior cannot be expressed as
a Gaussian, so the analytic expression for the posterior
discussed in \S\ref{sec:linearity} will generally have nonzero support for
negative albedo values. This is particularly problematic when the data is
not very constraining and the posterior is prior-dominated, as
positivity can be an \emph{extremely} informative prior
\citep[e.g.,][]{Fienup1978}.
We therefore recommend that in such cases
users of the \starry algorithm use MCMC to do inference, either
(1) sampling over the spherical
harmonic coefficients $\mathbf{y}$ and imposing a uniform prior in the range
$[0, 1]$ on the albedo values $\mathbf{a}$
evaluated on a discrete grid on the sphere or
(2) sampling over the albedo values $\mathbf{a}$ with the same uniform prior,
but using $\mathbf{y}$ to compute the actual light curve model. In both
cases, there exists a linear operator that transforms between
$\mathbf{y}$ and $\mathbf{a}$:
%
\begin{align}
    \label{eq:Py}
    \mathbf{a} = \mathbf{P} \mathbf{y}
\end{align}
%
and
%
\begin{align}
    \label{eq:PInvy}
    \mathbf{y} = \mathbf{P}^+ \mathbf{a}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:PInv}
    \mathbf{P}^+ =
    (\mathbf{P}^\top \mathbf{P} + \lambda \boldsymbol{\mathsf{I}})^{-1} \mathbf{P}^\top
\end{align}
%
is the pseudoinverse of $\mathbf{P}$, with (small) regularization parameter
$\lambda$ and where $\boldsymbol{\mathsf{I}}$ is the identity matrix.
%
Each row of the matrix $\mathbf{P}$ is constructed from the value
of each of the spherical harmonics at the corresponding point on the
grid; both $\mathbf{P}$ and its inverse may be pre-computed for
efficiency. In both cases, the grid should be fine enough to ensure
positivity over most of the sphere but not so fine as to throttle the
computation; as a rule of thumb, we find that grids with $2-5$ times
as many pixels
as spherical harmonic coefficients are sufficient.
%
The \href{https://rodluger.github.io/starry}{documentation}
includes tutorials on how to implement this in practice.

@@

The software presented in this work is open source under the MIT License and
is available at \url{https://github.com/rodluger/starry}, with documentation
and tutorials hosted at \url{https://rodluger.github.io/starry}.
The code used to generate the figures in this paper is hosted
at \url{https://github.com/rodluger/starrynight}.
% --- TODO ---
% with a permanent
% version archived at \url{https://doi.org/XX.YYYY/zenodo.ZZZZZZZ}.

\software{
    astroquery \citep{astroquery1,astroquery2},
    Eigen v3 \citep{eigen},
    exoplanet \citep{exoplanet},
    HEALPix \citep{healpix},
    pybind11 \citep{pybind11},
    pymc3 \citep{pymc3},
    scipy \citep{scipy},
    starry \citep{starry}.
}

>>>>

The present paper is an extension to the \starry algorithm,
adapting it to model phase curves and occultation light curves
in reflected light. The expressions
we derive are analytic: they may all be expressed in closed form
in terms of algebraic operations involving trigonometric functions
and (at times) elliptic integrals. We derive numerically
stable recursion relations for the efficient evaluation of all
expressions and code them within an autodifferentiation framework
to enable the computation of accurate derivatives for use in
gradient-based inference and optimization schemes.
Our code is fully
\href{https://github.com/rodluger/starry}{open-source},
comprehensively
\href{https://dev.azure.com/rodluger/starry/_test/analytics?definitionId=4}{unit-tested},
and supplemented with an extensive
\href{https://starry.readthedocs.io}{API documentation}
and suite of
\href{https://starry.readthedocs.io/en/latest/tutorials/}{tutorials}.
%
As in \citet{Luger2019}, in the caption of each of the figures
we provide links \codeicon\ to the exact \Python scripts that
generated them. Next to many of the equations we also provide
links \prooficon\ to \textsf{Jupyter} notebooks containing
detailed derivations and/or validations.

@@

The algorithm presented in this paper has been implemented in the \Python
package \starry, which can be installed from
\href{https://github.com/rodluger/starry}{GitHub} or via the Python package
manager \textsf{pip}. The algorithm is coded in a mixture of \textsf{C++}
with forward automatic differentiation using the \textsf{Eigen} library
\citep{eigen} and \Python with backward differentiation using
just-in-time compiled \textsf{theano}
operations \citep{theano}. The user interface, however, is purely in
\Python. The \textsf{theano} backend facilitates integration with the
\textsf{exoplanet} modeling package \citep{exoplanet} and in particular
with \textsf{pymc3} \citep{pymc3} for inference with gradient-based
Markov Chain Monte Carlo (MCMC) schemes
such as Hamiltonian Monte Carlo \citep[HMC;][]{Duane1987}
and No-U-Turn Sampling \citep[NUTS;][]{Hoffman2011}.
%
Complete \href{https://starry.readthedocs.io}{documentation} and an
extensive library of tutorials is available online.
The links next to each
of the figures \codeicon\, point to the \Python scripts used to generate
them and may also help in learning how to use \starry.

@@

It is important to note, however, that the \starry model is linear
only in the map coefficients $\mathbf{y}$. In any real application,
there will be uncertainty in the inputs of $\mathbf{X}$, such
as the orbital parameters, the occultor parameters, the scattering
law, etc. These parameters must typically be sampled over, since the model
is a nonlinear function of them. However, the analyticity---and in particular,
the differentiability---of the \starry model makes sampling via
gradient-based MCMC easy. Moreover, the linearity of the model
with respect to $\mathbf{y}$ allows one to efficiently marginalize over
those parameters when sampling over the nonlinear parameters.
Tutorials on how to do this can be found in the
\starry \href{https://starry.readthedocs.io}{documentation}.

@@

The third limitation concerns the flexibility of the \starry model.
While we presented ways to capture non-Lambertian scattering
in \starry (\S\ref{sec:nonlambertian}), there are certain aspects
of light curves in reflected light that cannot be captured by the model.
One example of this is shadowing. Craters on the moon or
volcanoes on Io can cast large shadows visible from space, particularly
if viewed near crescent phase. Unfortunately, there is no way to model
this within the \starry framework. Another example is multiple scattering,
as in optically thick atmospheres, for which a proper radiative transfer model
must be used. There is also the case of specular reflection,
or ``glint'', which is a pronounced signal for the Earth due its
oceans \citep{Robinson2014} and on Titan due to
its hydrocarbon lakes \citep{Barnes2011}. In principle, glint could be
modeled in the same way as non-Lambertian sccattering, by constructing the
linear operator $\mathbf{I}$ in such a way as to downweight portions of the
projected disk where $\vartheta_i \neq \vartheta_r$. In practice, however,
if the size of the glint spot is small (which is typically the case),
an expansion at extremely high $l$ ($l \sim 360$ in the case of the Earth)
would be required, which would not work for the reasons above.
Instead, it may be possible to combine the \starry algorithm with the
formalism of \citet{Haggard2018}, who derived analytic expressions for
phase curves of delta function maps ($\delta$-map), to model glint.
As discussed in \citet{LustigYaeger2018}, glint mapping is an extremely
powerful way to not only map terrestrial planets but also to confirm
their habitability via the presence of an ocean.  Lastly, the presence
of time-variable features, such as clouds, dust storms, or seasonal
variations of vegetation are not accounted for in the model, although
the \href{https://starry.readthedocs.io}{documentation} discusses
how one may approach the modeling of temporal features.

Finally, we would like to emphasize that while spherical harmonics are
a convenient basis for the purpose of computing light curves, they
have a significant drawback when it comes to modeling real planetary
surfaces: it can be difficult to strictly enforce physical
values of the albedo everywhere on the surface map when doing inference.
That is because there is no analytic way to determine whether a
spherical harmonic representation is positive-valued
(or restricted to a given range) everywhere on the sphere.
Instead, this must be checked numerically, by evaluating the function on a
discrete grid. This makes it somewhat cumbersome to implement positivity as a prior
when doing inference; in particular, this prior cannot be expressed as
a Gaussian, so the analytic expression for the posterior
discussed in \S\ref{sec:linearity} will generally have nonzero support for
negative albedo values. This is particularly problematic when the data is
not very constraining and the posterior is prior-dominated, as
positivity can be an \emph{extremely} informative prior
\citep[e.g.,][]{Fienup1978}.
We therefore recommend that in such cases
users of the \starry algorithm use MCMC to do inference, either
(1) sampling over the spherical
harmonic coefficients $\mathbf{y}$ and imposing a uniform prior in the range
$[0, 1]$ on the albedo values $\mathbf{a}$
evaluated on a discrete grid on the sphere or
(2) sampling over the albedo values $\mathbf{a}$ with the same uniform prior,
but using $\mathbf{y}$ to compute the actual light curve model. In both
cases, there exists a linear operator that transforms between
$\mathbf{y}$ and $\mathbf{a}$:
%
\begin{align}
    \label{eq:Py}
    \mathbf{a} = \mathbf{P} \mathbf{y}
\end{align}
%
and
%
\begin{align}
    \label{eq:PInvy}
    \mathbf{y} = \mathbf{P}^+ \mathbf{a}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:PInv}
    \mathbf{P}^+ =
    (\mathbf{P}^\top \mathbf{P} + \lambda \boldsymbol{\mathsf{I}})^{-1} \mathbf{P}^\top
\end{align}
%
is the pseudoinverse of $\mathbf{P}$, with (small) regularization parameter
$\lambda$ and where $\boldsymbol{\mathsf{I}}$ is the identity matrix.
%
Each row of the matrix $\mathbf{P}$ is constructed from the value
of each of the spherical harmonics at the corresponding point on the
grid; both $\mathbf{P}$ and its inverse may be pre-computed for
efficiency. In both cases, the grid should be fine enough to ensure
positivity over most of the sphere but not so fine as to throttle the
computation; as a rule of thumb, we find that grids with $2-5$ times
as many pixels
as spherical harmonic coefficients are sufficient.
%
The \href{https://starry.readthedocs.io}{documentation}
includes tutorials on how to implement this in practice.

@@

The software presented in this work is open source under the MIT License and
is available at \url{https://github.com/rodluger/starry}, with documentation
and tutorials hosted at \url{https://starry.readthedocs.io}.
The code used to generate the figures in this paper is hosted
at \url{https://github.com/rodluger/starrynight}.
% --- TODO ---
% with a permanent
% version archived at \url{https://doi.org/XX.YYYY/zenodo.ZZZZZZZ}.

\software{
    astroquery \citep{astroquery1,astroquery2},
    Eigen v3 \citep{eigen},
    exoplanet \citep{exoplanet},
    pybind11 \citep{pybind11},
    pymc3 \citep{pymc3},
    scipy \citep{scipy},
    starry \citep{starry}.
}

We would like to thank Nicolas Cowan, Christina Hedges, and the
Astronomical Data Group at the Center for Computational Astrophysics for
many thought-provoking discussions that helped shape this paper.
